---
title: "EV Charging Infrastructure Gaps"
author: "Sawyer Bringle & Liam Maumenee"
format:
  dashboard:
    orientation: rows
    theme: cosmo
    icon: github
server: shiny
---

```{r}
#| context: setup
#| include: false

# Packages
library(shiny)
library(leaflet)
library(tidyverse)
library(sf)
library(plotly)
library(scales)
library(DT)

load("clean_data/ev_final.RData")
load("clean_data/county_map.RData")
load("clean_data/stations_with_county.RData")
load("clean_data/desert_pool.RData")
load("clean_data/dashboard_summary.RData")
load("clean_data/predictive_model.RData")
load("clean_data/visualizations.RData")

# Drop stations missing coordinates for leaflet map
stations <- stations_with_county |>
  filter(!is.na(latitude), !is.na(longitude))

# Cosmetic palette + label maps used only by the dashboard
state_colors <- c("CO" = "#4A6FA5", "CT" = "#0072B2", "WA" = "#B7A57A")
state_labels <- c("CO" = "Colorado", "CT" = "Connecticut", "WA" = "Washington")

# Human-readable names for the Lasso predictors (used on the Predictive
# Model tab). Keys must match the variable names in predictive_model$coefs.
predictor_labels <- c(
  median_household_income = "Median Income",
  log_population          = "Log Population",
  pct_nonwhite            = "% Nonwhite",
  median_age              = "Median Age",
  evs_per_1000_pop        = "EVs / 1K Pop",
  stations_by_2020        = "Stations by 2020",
  stations_per_100k_pop   = "Stations / 100K Pop"
)

# Unpack summary stats so the existing chunks below don't have to change
n_counties_total    <- dashboard_summary$n_counties_total
total_evs           <- dashboard_summary$total_evs
total_stations_n    <- dashboard_summary$total_stations_n
avg_evs_per_station <- dashboard_summary$avg_evs_per_station
desert_threshold    <- dashboard_summary$desert_threshold
n_deserts           <- dashboard_summary$n_deserts
n_qualifying        <- dashboard_summary$n_qualifying
desert_state_counts <- dashboard_summary$desert_state_counts
```

# Overview

## Row {height="22%"}

```{r}
#| content: valuebox
#| title: "Counties Studied"
#| color: "#4A6FA5"
#| align: "center"
list(value = format(n_counties_total))
```

```{r}
#| content: valuebox
#| title: "Total EVs on Road"
#| color: "#B7A57A"
#| align: "center"
list(value = format(total_evs, big.mark = ","))
```

```{r}
#| content: valuebox
#| title: "Charging Stations"
#| color: "#4A6FA5"
#| align: "center"
list(value = format(total_stations_n, big.mark = ","))
```

```{r}
#| content: valuebox
#| title: "Avg EVs per Station"
#| color: "#B7A57A"
#| align: "center"
list(value = format(avg_evs_per_station, big.mark = ","))
```

## Row {height="78%"}

### Column {width="65%"}

```{r}
#| title: "Summary by State"
# Built upstream in 01_data_cleaning.R as dashboard_summary$state_summary.
# The dashboard just renders it — no recomputation here.
knitr::kable(dashboard_summary$state_summary, align = "cccccccc")
```


### Column {width="35%"}

::: {.card title="About this Dashboard"}
**Project:** ECNS 460 Term Project — *EV Charging Infrastructure Gaps*

**Authors:** Sawyer Bringle & Liam Maumenee

**Montana State University**

**Course:** ECNS 460

**Question:** Where are the gaps between EV adoption and charging infrastructure, and which counties are most at risk of becoming charging deserts?

:::

# Visualizations

## Row {.tabset}

```{r}
# pull chart from visualizations Rdata set for each figure 1-5
#| title: "Top Counties by EVs"
visualizations$fig1
```

```{r}
#| title: "Income vs EV Adoption"
visualizations$fig2
```

```{r}
#| title: "Stations vs EVs"
visualizations$fig3
```

```{r}
#| title: "Percent Nonwhite Population vs. Charging Access"

visualizations$p_co
visualizations$p_ct
visualizations$p_wa

```

```{r}
#| title: "Top Charging Deserts"
visualizations$fig5
```




# Charging Deserts

## Row {height="22%"}

```{r}
#| content: valuebox
#| title: "Charging-Desert Counties"
#| color: "#4A6FA5"
#| align: "center"
list(value = n_deserts)
```

```{r}
#| content: valuebox
#| title: "Threshold (EVs / Station)"
#| color: "#B7A57A"
list(value = format(round(desert_threshold, 1), big.mark = ","))
```

```{r}
#| content: valuebox
#| title: "Qualifying Counties"
#| color: "#4A6FA5"
#| align: "center"
list(value = n_qualifying)
```



## Row {height="78%"}

### Column {width="40%"}

::: {.card title="What is a 'Charging Desert'?"}

A **charging desert** is a county where existing EV charging infrastructure is most stretched relative to the EV fleet already on the road.

**Our definition:** A county in the top quartile of EVs-per-station, restricted to counties with at least 50 EVs and at least one station (so the ratio is well-defined).

**Why this matters:** Counties with high EVs-per-station ratios face longer wait times, more competition for chargers, and a weaker case for new EV buyers — which can stall further adoption.

**Charging Desert Threshold:** We rank every qualifying county by its EVs-per-station ratio and take the 75th percentile as the cut-off. Any county at or above that line is flagged as a desert. We chose a quartile-based threshold instead of a fixed number because the data is heavily right-skewed and a fixed number would either flag almost every county or almost none, depending on where it's set. A quartile is also relative, so the definition stays meaningful as adoption grows. With our current data the cut-off lands at roughly **70 EV's Per Station.**

**Key Findings:**

- 22 out of 87 qualifying ones meet the desert threshold — concentrated in WA (14) and CO (7), with only 1 in CT.
- Deserts are mostly large suburban / exurban counties, not rural ones. Places like Snohomish (WA), Douglas and El Paso Counties (CO), where EV adoption has outpaced station rollout.
- The gap between desert and non-desert counties is large: the median ratio is roughly 118 EVs/station in deserts vs. 26 in non-deserts. This is about a 4.5× difference.
- Deserts are not a low-income situation: median household income is actually higher in desert counties (\~94K) than non-desert counties (\~74K). The gap is driven by demand outrunning supply in commuter counties, not by underinvestment in poor areas. This makes sense, as higher-income areas are more likely to have more EV's as they are on the higher end of the price spectrum for car brands.

:::

### Column {width="60%"}

```{r}
#| title: "Top 10 Charging Desert Counties"
desert_pool |>
  arrange(desc(evs_per_station)) |>
  head(10) |>
  transmute(
    Rank             = row_number(),
    State            = recode(state, !!!state_labels),
    County           = county,
    `EVs on Road`    = comma(evs_on_road),
    Stations         = total_stations,
    `EVs / Station`  = round(evs_per_station, 1),
    Population       = comma(total_population),
    `Median Income`  = paste0("$", comma(median_household_income))
  ) |>
  DT::datatable(
    rownames = FALSE,
    options = list(
      dom = "t",
      pageLength = 10,
      scrollY = "100%",
      autoWidth = TRUE
    )
  )
```

# Interactive Map

##  {.sidebar width="320px"}

```{r}
radioButtons(
  "map_state", "Select State:",
  choices = c("Colorado"    = "CO",
              "Connecticut" = "CT",
              "Washington"  = "WA"),
  selected = "CO",
  inline = TRUE
)

selectInput(
  "fill_var", "Color Counties By:",
  choices = c(
    "EVs per Station"             = "evs_per_station",
    "EVs per 1,000 Residents"     = "evs_per_1000_pop",
    "Stations per 100K Residents" = "stations_per_100k_pop",
    "Total EVs on Road"           = "evs_on_road",
    "Total Charging Stations"     = "total_stations",
    "Median Household Income"     = "median_household_income",
    "Median Age"                  = "median_age",
    "Percent Nonwhite"            = "pct_nonwhite",
    "Percent Hispanic"            = "pct_hispanic",
    "Total Population"            = "total_population"
  ),
  selected = "evs_per_station"
)

checkboxInput("show_stations", "Show Station Markers", value = TRUE)

tags$hr()
tags$h5("Summary")
textOutput("map_station_count")
tags$br()
textOutput("map_ev_count")


```

## Map

```{r}
#| title: "Stations & County Choropleth"
leafletOutput("station_map")
```

```{r}
#| context: server

# Constants
state_bbox <- list(
  CO = list(lng = -105.5, lat = 39.0, zoom = 7),
  CT = list(lng =  -72.7, lat = 41.6, zoom = 9),
  WA = list(lng = -120.5, lat = 47.4, zoom = 7)
)

# match variables to label names for sidebar selection
var_label <- function(v) {
  switch(v,
    "evs_per_station"          = "EVs per Station",
    "evs_per_1000_pop"         = "EVs per 1K Pop.",
    "stations_per_100k_pop"    = "Stations per 100K Pop.",
    "evs_on_road"              = "Total EVs on Road",
    "total_stations"           = "Total Stations",
    "median_household_income"  = "Median Income ($)",
    "median_age"               = "Median Age",
    "pct_nonwhite"             = "% Nonwhite",
    "pct_hispanic"             = "% Hispanic",
    "total_population"         = "Total Population",
    v
  )
}

# Reactive filters: a function that re-runs automatically whenever the sidebar input changes
# Both filters depend on input$map_state so if the user clicks a different state or variable object,
# the objects below are recomputed and every chunk that calls them gets fresh data. 
# Downstream code calls them like functions, not regular variables
# This took us a while to figure out, we originally had it formatted inside of the sidebar
# which didn't allow for updates automatically.

filtered_counties <- reactive({
  county_map |> filter(state_abbr == input$map_state)
})

filtered_stations <- reactive({
  stations |> filter(state == input$map_state)
})


# Sidebar summary text
output$map_station_count <- renderText({
  paste("Stations shown:", format(nrow(filtered_stations()), big.mark = ","))
})

output$map_ev_count <- renderText({
  total <- sum(filtered_counties()$evs_on_road, na.rm = TRUE)
  paste("Total EVs on road:", format(total, big.mark = ","))
})


# Base Leaflet map, drawn once, CO is the default state
output$station_map <- renderLeaflet({
  leaflet() |>
    addProviderTiles(providers$CartoDB.Positron) |>
    setView(state_bbox$CO$lng, state_bbox$CO$lat, state_bbox$CO$zoom)
})


# Fly to selected state when selected on sidebar
# With only three states being used, the other 47 states are still shown on the map, but very useless to us.
# The observe() function allows us to zoom in on one of the three states, instead of viewing the whole U.S. 
observe({
  b <- state_bbox[[input$map_state]]
  leafletProxy("station_map") |>
    flyTo(b$lng, b$lat, b$zoom)
})


# Update choropleth + popup whenever the state or color-variable changes.
# observe() also re-runs when its inputs change, but instead of returning a value, it redraws the map. 
# Then we use leafletProxy() instead of redrawing the whole map for efficiency
# clearShapes() and clearControls() wipe the old polygons and legend so we can add the new ones
# without the layers piling up on every change.
observe({
  counties <- filtered_counties() |>
    filter(!is.na(.data[[input$fill_var]]))
  fv       <- input$fill_var
  vals     <- counties[[fv]]
  pal      <- colorNumeric("YlOrRd", domain = vals, na.color = "gray80")

  popup_text <- paste0(
    "<b>", counties$county_clean, ", ", counties$state_abbr, "</b><br>",
    "EVs on Road: ",     format(counties$evs_on_road, big.mark = ","), "<br>",
    "Stations: ",        counties$total_stations, "<br>",
    "EVs per Station: ", round(counties$evs_per_station, 1), "<br>",
    "Population: ",      format(counties$total_population, big.mark = ","), "<br>",
    "Median Income: $",  format(counties$median_household_income, big.mark = ",")
  )

  leafletProxy("station_map") |>
    clearShapes() |>
    clearControls() |>
    addPolygons(
      data        = counties,
      fillColor   = ~pal(counties[[fv]]),
      fillOpacity = 0.65,
      color       = "white",
      weight      = 1,
      popup       = popup_text,
      highlightOptions = highlightOptions(
        weight = 2, color = "#666",
        fillOpacity = 0.85, bringToFront = TRUE
      )
    ) |>
    addLegend(
      position = "bottomright",
      pal      = pal,
      values   = vals,
      title    = var_label(fv),
      opacity  = 0.8
    )
})

# Station markers: separate observer so toggling "Show Station Markers" only
# redraws the dots, not the choropleth underneath.
#
# Like the choropleth observer, this re-runs whenever any input it reads
# changes. It depends on input$show_stations and on filtered_stations()
# (which depends on input$map_state). clearMarkers() wipes the old dots
# every time, then we add fresh ones if the checkbox is on.
observe({
  leafletProxy("station_map") |> clearMarkers()

  if (input$show_stations) {
    leafletProxy("station_map") |>
      addCircleMarkers(
        data        = filtered_stations(),
        lat         = ~latitude,
        lng         = ~longitude,
        radius      = 4,
        color       = "#2166AC",
        fillOpacity = 0.70,
        stroke      = FALSE
      )
  }
})
```



# Predictive Model

## Row {height="100%"}

### Column {width="35%"}

::: {.card title="About this Model"}

Test R-Squared = 0.16 | Test RMSE = 2.465 | 86 Counties

We built a Lasso regression to complement the descriptive findings and identify which counties are on the steepest charging-growth trajectories.

**Outcome:** Because we did not have time-series EV registration data, we use proportional growth in station count between 2020 and 2024 as the outcome variable. Stations follow demand reasonably closely, so this is a defensible proxy for where EV adoption is accelerating. The model is, in effect, asking: holding current EV stock and 2020 infrastructure fixed, which counties are seeing the fastest expansion of charging supply?

**Why Lasso:** We chose Lasso over a more flexible model like random forest for two reasons. First, Lasso produces a sparse set of coefficients that can be interpreted directly and easily. Second, we covered Lasso in class and have a stronger working understanding of it than of less-familiar alternatives.

**Predictors:** Median household income, log population, percent nonwhite, median age, EVs per 1,000 residents, stations open by 2020, stations per 100,000 residents, and an early-adopter indicator.

**Tuning:** Penalty parameter tuned by 5-fold cross-validation on a 75/25 train-test split, then evaluated once on the held-out test set.

**What the coefficients say:** The retained coefficients run in the expected directions. Counties with more stations already in place by 2020 see slower proportional growth than counties with fewer; counties with higher current per-capita EV adoption see faster station expansion. Demographic predictors enter with smaller coefficients, indicating that 2020 infrastructure and current adoption are doing most of the predictive work.

**Assumptions:**

- Station growth is a proxy for EV growth, not a direct measure.
- Counties with zero stations in 2020 are dropped (growth rate undefined), so we miss counties going from 0 to 1+ stations.
- A three-state sample is small, which is part of why we leaned on Lasso's regularization rather than a more flexible model.

:::


### Column {width="40%"}

```{r}
#| title: "Lasso Coefficients (predictors normalized)"
predictive_model$coefs |>
  filter(term != "(Intercept)") |>
  mutate(
    term = unname(predictor_labels[term]),
    sign = if_else(estimate > 0, "Positive", "Negative")
  ) |>
  plot_ly(
    x      = ~estimate,
    y      = ~reorder(term, estimate),
    color  = ~sign,
    colors = c("Positive" = "#0072B2", "Negative" = "#F45D40"),
    type   = "bar",
    orientation = "h"
  ) |>
  layout(
    xaxis = list(title = "Standardized coefficient", zerolinecolor = "#888"),
    yaxis = list(title = ""),
    showlegend = FALSE,
    margin = list(l = 30, r = 20, t = 20, b = 50)
  )
```

### Column {width="25%"}

```{r}
#| title: "Top 20 Counties by Predicted Growth"
predictive_model$scored |>
  arrange(desc(predicted_growth)) |>
  slice_head(n = 20) |>
  transmute(
    Rank           = row_number(),
    County         = paste0(county, ", ", state),
    `Pred. Growth` = percent(predicted_growth, accuracy = 1)
  ) |>
  knitr::kable(align = "clr")
```




# Summary

## Column {width="48%"}

::: {.card title="Key Findings"}

The gap is not where you'd expect. Across Colorado, Connecticut, and Washington, the counties where EV adoption has run furthest ahead of charging build-out are predominantly suburban and exurban commuter belts. Places like Snohomish (WA), Douglas (CO), and El Paso (CO). 

Using the 75th percentile of qualifying counties, a rough cutoff of 70 EVs per station gave us a great baseline. 22 of 87 qualifying counties cross that line, concentrated in WA (14) and CO (7), with only 1 in CT. Deserts hold a disproportionate share of the EV fleet. The 20% of counties flagged as deserts hold roughly 71% of all EVs across the three states. The median desert county has ~118 EVs/station versus ~26 in non-desert counties — about a 4.5x difference. 

Lasso further confirms the infrastructure gap. With a test R-Squared of 0.162, the Lasso shows that 2020 infrastructure stock and current per-capita adoption dominate the prediction of where new stations grow next, while demographic predictors enter with smaller coefficients than we hypothesized in Stage 2.

The interactive map is a great tool that state planners, EV station developers, and anyone curious about where EV stations are placed, or if their county is at 
risk for longer wait-times at charging stations.

:::


### Column {width="52%"}

::: {.card title="Data Sources & Tools"}

**Data**

- **AFDC Alternative Fuel Stations** — U.S. Department of Energy, Alternative Fuels Data Center. Charging-station locations, ports, and status.
- **EV Registrations** — Colorado Energy Office (county-level), CT DMV (vehicle-level), WA Department of Licensing (vehicle-level).
- **ACS 5-Year Estimates (2023)** — U.S. Census Bureau. Tables DP05 (population and demographics) and S1901 (median household income).
- **TIGER/Line Shapefiles** — U.S. Census Bureau, accessed via the `tigris` R package. County polygons for the spatial join and choropleth.

**Tools**

R | tidyverse | sf | tigris | janitor | tidymodels (glmnet) | leaflet | plotly | DT | Quarto | Shiny

**References**

*Full citations to be added in the project report.*

:::

::: {.card title="About"}
**Course** \| ECNS 460 — Spring 2026

**Instructor** \| Nick Hagerty

**Authors** \| Sawyer Bringle & Liam Maumenee

**Audience** \| State transportation planners, EV charging operators, and policymakers — the stakeholders who decide where the next round of charging investment goes.

**Source code** \| `01_data_cleaning.R` through `04-dashboard.qmd` in the project repository.
:::
